Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-01-28 paper/main.tex
updates
Updates

\section{Introduction}
Understanding cause-and-effect relationships between variables is a fundamental
objective in many scientific fields. These relationships reveal the mechanisms
behind observed phenomena and guide effective interventions or policy
decisions. \emph{Causal Discovery} methods aim to identify such relationships
among random variables using observational data. Approaches to causal discovery
have been developed within both the Directed Acyclic Graphs (DAGs) and
Structural Equation Models (SEMs) frameworks, each adapting a different approach.

In the DAG literature, the primary focus has been on developing automated
algorithms to learn causal structures from datasets. These efforts have led to
numerous causal discovery algorithms, including constraint-based methods like
the PC algorithm \citep{Spirtes2001} and Fast Causal Inference (FCI)
\citep{Spirtes2000}), score-based methods such as Hill-Climb Search and Greedy
Equivalence Search (GES) \citep{Chickering2002}, and continuous
optimization-based methods like NO TEARS \citep{Zheng2018} and DAGMA
\citep{Bello2022}. In contrast, the SEM literature has focused on developing
tools to assist researchers in manually constructing or modifying models,
enabling experts to incorporate their domain knowledge in the model building
process. Researchers typically begin with an initial model based on their
expertise and then use these tools to guide modifications that improve the
model's fit to data. This process is commonly known as \emph{Specification
Search} \citep{Long1983} and uses method such as modification indices, and
Wald-based tests \citep{Marcoulides2018}.

@@

Figure~\ref{fig:intro} shows an example to highlight some of these issues using 
Adult income dataset.

These challenges highlight the need for methods that effectively combine
automated tools with expert-driven model refinement.  As a result of these
issues, expert knowledge integration is vital to get usable results from these
algorithms. One way to do this to specify expert knowledge in terms of
fixing edges, blacklisting edges, or specifying tiered knowledge.
The other ways is to do post hoc modification to the learned model using
implied CI based model testing \citep{Ankan2021, Ankan2023}. In this testing
procedure, users test whether the implied CIs of the DAG hold in the dataset or
not using statistical CI tests. Based on the results of these statistical
tests, we can accordingly make modifications to the current model to fix the
violations. However, this CI based testing also has a few drawbacks: 1) CIs are
implied only by missing edges in the model, so we can not test for already
existing edges. 2) In high sample size scenarios, p-values can suggest a
significant association between variables even if they have a very weak
relationship. 3) Number of implied CIs can be large making it difficult to
manually go through each of them; for example, in the moderately sized alarm
network with $ 37 $ node, there are $287$ implied CIs.

There has been work on trying to come up with a single test to combine all
these test to give a single answer \citep{Eulig2023, Shipley2000}. However,
what is missing is the guide on how to how to fix these tests. Additionally
there has been a lot of work to integrate expert knowledge into causal
discovery algorithms which has shown significant improvement. To tackle these
issues with CI based testing, we take inspiration from \emph{modification
indices} in specification search. The modification indices method tries to
identify and rank changes in the model such that modifications that improve the
model fit the most are ranked higher.This ranking allows researchers to focus
first on the suggestions that can improve the model the most and while
integrating their expertise to choose the most optimal modifications. Analogous
to this, in this paper we develop a measure for partial association for mixed
data that allows us to rank the CI violations. Additionally, this measure of
conditional association also allows us tackle other issues with CI based
testing outlined in the previous paragraphs.

@@

			% \begin{scope}[xshift=-3cm]
			% 	\draw[thick, ->, double, cyan] (2.5,0) -- (3,0);
			% 	\node[rectangle, align=center, inner sep=1pt] at (5, 0) {
			% 		\begin{minipage}{.2\textwidth}
			% 			\begin{equation*}
			% 				\begin{split}
			% 					\textnormal{Age} &\ci \textnormal{Incm} \rvert \textnormal{Edct, Hrpw, Sex} \\
			% 					\textnormal{Age} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
			% 					\textnormal{Age} &\ci \textnormal{Sex} \\
			% 					\textnormal{Hrpw} &\ci \textnormal{Occp} \rvert \textnormal{Edct, Sex} \\
			% 				\end{split}
			% 			\end{equation*}
			% 		\end{minipage}
			% 		};
			% \end{scope}

@@

\subsection{Ranking using Correlation Coefficients}

\subsection{Measures of (Conditional) Association}
\todo[inline]{Complete this section}
As our DAG construction procedure relies on a measure of conditional
association between variables, in this section we list out some of the already
known measures for specific data types. These measure of associations are
typically the effect size of various conditional indepdnence tests but they
don't need to be. However, there are no generalized measure of association for
mixed data, that we propose in the next section. This is by no means an
exhaustive list, rather just to give an idea of what can be done.

@@

\paragraph{$ X $ is ordinal and $ Y $ is continuous}
Polyserial Correlation

When $ \bm{Z} = \emptyset $, ordinal variable can be assumed to be coming from
a thresholded normal distribution, estimate the thresholds and latent variable
to compute the correlation coefficient.

\paragraph{$ X $ and $ Y $ are ordinal}
Polychoric Correlation

>>>>

\section{Introduction}
Understanding cause-and-effect relationships between variables is a fundamental
objective in many scientific fields. These relationships reveal the mechanisms
behind observed phenomena and guide effective interventions or policy
decisions. \emph{Causal Discovery} methods aim to discover such relationships
among random variables using observational data. Approaches to causal discovery
have been developed within both the Directed Acyclic Graphs (DAGs) and
Structural Equation Models (SEMs) frameworks, each adapting a different
approach.

In the DAG literature, the primary focus has been on developing automated
algorithms to learn causal structures from datasets. These efforts have led to
numerous causal discovery algorithms, including constraint-based methods like
the PC algorithm \citep{Spirtes2001} and Fast Causal Inference (FCI)
\citep{Spirtes2000}), score-based methods such as Hill-Climb Search and Greedy
Equivalence Search (GES) \citep{Chickering2002}, and continuous
optimization-based methods like NO TEARS \citep{Zheng2018} and DAGMA
\citep{Bello2022}. In contrast, the SEM literature has focused on developing
tools to assist researchers in manually constructing or modifying models,
enabling experts to incorporate their domain knowledge in the model building
process. Researchers typically begin with an initial model based on their
expertise and then use these tools to guide modifications that improve the
model's fit to data. This process is commonly known as \emph{Specification
Search} \citep{Long1983} and uses method such as modification indices, and
Wald-based tests \citep{Marcoulides2018}. While there has been some effort in
DAG literature as well towards allowing for expert knowledge but these methods
focus towards providing the expert knowledge to the automated algorithms
\citep{Bang2023, Meek1995}. \todo[inline]{Whether to keep this sentence?}

@@

Figure~\ref{fig:intro} shows an example to highlight some of these issues using
the Adult income dataset \citep{Becker1996}. As can be seen in the figure
depending on the sample size and the algorithm used, we get vastly different
results.

The other ways is to do post hoc modification to the learned model using
implied CI based model testing \citep{Ankan2021, Ankan2023}. In this testing
procedure, users test whether the implied CIs of the DAG hold in the dataset or
not using statistical CI tests. Based on the results of these statistical
tests, we can accordingly make modifications to the current model to fix the
violations. However, this CI based testing also has a few drawbacks: 1) CIs are
implied only by missing edges in the model, so we can not test for already
existing edges. 2) In high sample size scenarios, p-values can suggest a
significant association between variables even if they have a very weak
relationship. 3) Number of implied CIs can be large making it difficult to
manually go through each of them; for example, in the moderately sized alarm
network with $ 37 $ node, there are $287$ implied CIs.

There has been work on integrating expert knowledge into model building process
and has been quite successful such as tpc and background knowledge
specification. There has been work on trying to come up with a single test to
combine all these test to give a single answer \citep{Eulig2023, Shipley2000}.
However, what is missing is the guide on how to how to fix these tests.
Additionally there has been a lot of work to integrate expert knowledge into
causal discovery algorithms which has shown significant improvement. To tackle
these issues with CI based testing, we take inspiration from \emph{modification
indices} in specification search. The modification indices method tries to
identify and rank changes in the model such that modifications that improve the
model fit the most are ranked higher.This ranking allows researchers to focus
first on the suggestions that can improve the model the most and while
integrating their expertise to choose the most optimal modifications. Analogous
to this, in this paper we develop a measure for partial association for mixed
data that allows us to rank the CI violations. Additionally, this measure of
conditional association also allows us tackle other issues with CI based
testing outlined in the previous paragraphs.

@@

@@

\subsection{Measures of Conditional Association}
As our DAG construction procedure relies on a measure of conditional
association between variables, in this section we list out some of the already
known measures for specific data types. These measure of associations are
typically the effect size of various conditional independence tests but they
don't need to be. However, there are no generalized measure of association for
mixed data, that we propose in the next section. This is by no means an
exhaustive list, rather just to give an idea of what can be done.

@@

\paragraph{$ X $ is ordinal and $ Y $ is continuous or ordinal}
Polyserial Correlation (when one is oridinal and the other is continuous) and
polychoric correlation (when both are ordinal) can be use. Here the assumption
is made that $ X $ is a thresholded version of an underlying continuous normal
variable. The method then tries to learn the thresholds for this latent
variable while trying to maximize the correlation between $ X_{lat} $ and $ Y
$. This gives us values that are Pearson's correlation coefficient.
